Local Spectral Clustering of Density Upper Level Sets

### 2019-05-22 neurips/old_versions/local_spectral_density_clustering_tuesday.tex
wednesday edits, reorganized folder

\newcommand{\diam}{\rho}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
% \newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\Phi_{u}}
\newcommand{\Psibf}{\Psi_{u}}
\newcommand{\taubf}{\tau_{u}}
\newcommand{\dist}{\mathrm{dist}}

@@

\begin{abstract}
\vskip 0.1 in
  Spectral clustering methods are a family of popular nonparametric clustering
  tools.  Recent works have proposed and analyzed \emph{local} spectral methods, such as Personalized PageRank (\ppr),
  which extract clusters using locally-biased random walks around a user-specified
  seed node.  In contrast to existing results, we analyze PPR in a traditional
  statistical learning setup, where we obtain samples from an unknown
  distribution, and aim to identify connected regions of high-density (density
  clusters).  We prove that PPR, run on a neighborhood graph, extracts
  sufficiently salient density clusters, and provide empirical support for our theory.
\end{abstract}

@@

Local spectral clustering techniques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (like
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of a local clustering algorithm on $\Xbf$ reveals
about the unknown density $f$.  In particular, we examine the ability of the \pprspace algorithm to recover \emph{density clusters} of $f$, which are defined as the
connected components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$
for some threshold $\lambda > 0$ (a central object of interest in the
classical statistical literature on clustering, dating back to
\citet{hartigan1981}).

\paragraph{PPR on a neighborhood graph} We now describe the clustering algorithm that will be our focus for the rest of 
the paper. We start with the geometric graph that we form based on the samples 
$\Xbf$: for a radius $r > 0$, we consider the \emph{$r$-neighborhood graph} of 
$\Xbf$, denoted $G_{n,r}=(V,E)$, an unweighted, undirected graph with vertices
$V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i - x_j}
\leq  r$, where $\norm{\cdot}$ denotes Euclidean norm. We denote by $\Abf \in \Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if $(u,v) \in E$ and $0$ otherwise, by $\Dbf$ the diagonal degree matrix, with $\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$, and by $\Ibf{}$ the $n \times n$ identity matrix.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, with respect to  
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the lazy random walk matrix over $G_{n,r}$ 
and $e_{v}$ denotes the indicator vector for node $v$ (with a 1 in the $v$th
position and 0 elsewhere).

For a level $\beta > 0$ and a target volume $\vol_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \set{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}}.
\end{equation}
We will use normalized cut to determine which sweep cut $S_{\beta}$ is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V \setminus S$, we define the cut as $\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c} \Abf_{uv}$, the volume as $\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$, and the \emph{normalized cut} as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}, we output the cluster estimate $\widehat{C} = S_{\beta^*}$ which has minimum normalized cut $\Phi(S_{\beta^{\star}}; G_{n,r})$. For concreteness, we summarize this procedure in Algorithm \ref{alg: ppr}.

@@

\begin{definition}[Symmetric set difference]
	\label{def: symmetric_set_diff}
	For an estimator \smash{$\Cest \subseteq \Xbf$} and set
	$\mathcal{S} \subseteq \Reals^d$, the symmetric set difference of $\Cest$ and $\mathcal{S} \cap \Xbf = \mathcal{S}[\Xbf]$ is 
	\begin{equation}
	\label{eqn: misclassification_rate}
	\Delta(\Cest, \mathcal{S}) := \abs{\Cest \setminus \mathcal{S}[\Xbf] \cup \mathcal{S}[\Xbf] \setminus \Cest}.
	\end{equation}
\end{definition}

@@

\begin{enumerate}
	\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we introduce a set of natural geometric conditions on the density cluster $\Cset$, formalize a measure of difficulty based on these geometric conditions, and show that when Algorithm \ref{alg: ppr} is properly initialized, the symmetric set difference of $\widehat{C}$ and a thickened version of the density cluster $\Csig$ is upper bounded by this difficulty measure.

	\item We further show that if the density cluster $\Cset$ is particularly well-conditioned, Algorithm \ref{alg: ppr} will consistently estimate a density cluster in the sense of \eqref{eqn: consistent_density_cluster_recovery}.

	\item In Section \ref{sec: analysis}, we detail some of the analysis required to prove our main results, and expose the part various geometric quantities play in the ultimate difficulty of the clustering problem.

	\item In Section \ref{sec: experiments}, we empirically
	demonstrate the tightness of our analysis, and provide examples showing how violations of the geometric conditions we require manifestly
	impact density cluster recovery by \ppr.  
\end{enumerate}

@@

\paragraph{Related work}

In addition to the background given previously, a few related lines of work are worth
highlighting. Similar in spirit to our results are the works
\citep{shi2009,schiebinger2015}, who study the consistency of
spectral algorithms in recovering the latent labels in certain
nonparametric mixture models. These results focus on global rather than local
algorithms, and as such impose global rather than local conditions on the nature
of the density. Moreover, they do not in general guarantee recovery of density
clusters, which is the focus in our work. Perhaps most importantly, these works rely on general cluster saliency conditions, which implicitly depend on many distinct geometric aspects of the cluster $\Cset$ under consideration. We make this dependence more explicit, and in doing so expose the role each geometric condition plays in the clustering problem.

Additionally, we note that density clustering and level set estimation is a well-studied problem. \citep{polonik1995, rigollet2009} study density clustering under symmetric set difference, \citep{tsybakov1997, singh2009} exhibit minimax optimal level set estimators under Hausdorff loss and \citep{hartigan1981, chaudhuri2010} consider consistent estimation of the cluster tree, to note but a few works on the subject. Our goal is not to improve on these results, or offer yet another algorithm for level set estimation; indeed, seen as a density clustering algorithm, \pprspace has none of the optimality guarantees of the previous works. Rather, our motivation is to better understand and characterize the distinctions between those density clusters which are well conditioned for local spectral algorithms, and those which are not.

@@

As mentioned previously, successful recovery of a density cluster by \pprspace requires the density cluster to be geometrically well-conditioned. At a minimum, we wish to avoid sets $\Cset$ which contain arbitrarily thin bridges or spikes, and therefore as in \cite{chaudhuri2010} we introduce a buffer zone around $\Cset$. Let $B(x,r)$ be the closed ball of radius $r > 0$ centered at $x \in \Rd$. For $\lambda > 0$, consider a given cluster $\Cset \in \Cbb_f(\lambda)$. We denote the distance between $x$ and $\Cset$ as $\dist(x,\Cset) := \inf_{y \in \Cset} \norm{y - x}$, and for a given $\sigma > 0$, we refer to $\Csig := \set{x \in \Reals^d: \dist(x,\Cset) \leq \sigma}$ as the $\sigma$-expansion of $\Cset$. We now state our conditions with respect to $\Csig$.

@@

	\item 
	\label{asmp: low_noise_density}
	\emph{Low noise density:} There exists $\gamma,c_0 > 0$ such that for all $x
	\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
	$$
	\inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma}.
	$$

@@

To motivate these conditions, for an arbitrary graph $G = (V,E)$ and subset of vertices $S \subseteq V$, consider the normalized cut $\Phi(S;G)$ (defined as in \eqref{eqn: normalized_cut}), as well as the mixing time of a random walk over the induced subgraph $G[S]$ (defined in Section \ref{sec: analysis} by \eqref{eqn: mixing_time}, and denoted by $\tau_{\infty}(G[S])$). In \citet{zhu2013}, it is shown that for a constant $c > 0$, the \pprspace estimate $\widehat{C}$ of $S$ satisfies
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\widehat{C} \setminus S; G) + \vol(S \setminus \widehat{C}; G) \leq c \bigl( \Phi(S,G) \cdot \tau_{\infty}(G[S])\bigr) \vol(S;G)
\end{equation}
where the left hand side resembles a (degree-weighted) form of the symmetric set difference of \eqref{eqn: misclassification_rate}.

As we will formally show in Section \ref{sec: analysis}, the conditions \ref{asmp: bounded_density}-\ref{asmp: bounded_volume} allow us to upper bound the normalized cut $\Phi(\Csig[\Xbf]; G_{n,r})$, and the mixing time $\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$. Intuitively, the low noise \ref{asmp: cluster_separation} and cluster separation \ref{asmp: low_noise_density} assumptions yield an upper bound on $\cut(\Csig[\Xbf]; G_{n,r})$, the lower bound on density in \ref{asmp: bounded_density} yields a lower bound on $\vol(\Csig[\Xbf]; G_{n,r})$, and along with \ref{asmp: bounded_volume}, which ensures $\vol(\Csig[\Xbf]; G_{n,r}) \leq \vol(\Csig[\Xbf]^c; G_{n,r})$, these imply an upper bound on the normalized cut. \ref{asmp: bounded_density} and \ref{asmp: embedding} preclude bottlenecks in the induced subgraph $G_{n,r}[\Csig[\Xbf]]$, and combined with the upper bound on diameter in \ref{asmp: embedding}, they yield an upper bound on the mixing time over this subgraph.

\paragraph{Condition number} We will return to the topic of conditions in Section \ref{sec: analysis}. Now, we define the condition number, $\kappa(\Cset)$, which reflects the difficulty of the local spectral clustering task. Motivated by \eqref{eqn: graph_symmetric_set_difference_1}, we will set $\kappa(\Cset)$ to be an upper bound on $\Phi(\Csig[\Xbf]; G_{n,r}) \cdot \tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$. The smaller $\kappa(\Cset)$ is, the more success \pprspace will have in recovering $\Cset$. Let $\theta := (r, \sigma, \lambda, \lambda_{\sigma}, \Lambda_{\sigma}, \gamma, \diam, L)$ contain those geometric parameters detailed in \ref{asmp: bounded_density} - \ref{asmp: bounded_volume}.

\begin{definition}[Well-conditioned density clusters]
	For $\lambda > 0$ and $\Cset \in \Cbb_f(\lambda)$, let $\Cset$ satisfy \ref{asmp: bounded_density} - \ref{asmp: bounded_volume} for some $\theta$. Then, for universal constants $c_1, c_2, c_3 > 0$ to be specified later, we set
	\begin{equation}
	\label{eqn: condition_number_1}
	\Phibf(\theta) 
	:= c_1 r \frac{d}{\sigma} \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} - c_0 \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}},~ 
	\taubf(\theta) := c_2 \frac{\Lambda_{\sigma}^4 d^3 \rho^2 L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{1}{r}\right) + c_3
	\end{equation}
	and letting $\kappa(\Cset) := \Phi_{u}(\theta) \cdot \tau_{u}(\theta)$, we call $\Cset$ a \textrm{$\kappa$-well-conditioned density cluster}.
\end{definition}

We note that $\Phibf(\theta)$ and $\taubf(\theta)$ are exactly the upper bounds on $\Phi(\Csig[\Xbf]; G_{n,r})$ and $\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$ that we derive in Section \ref{sec: analysis}.

@@

As is typical in the local clustering literature, our algorithmic results will be stated with respect to specific ranges of each of the user-specified parameters.

In particular, for a well-conditioned density cluster $\Cset$ (with respect to some $\theta$), we require
\begin{align}
\label{eqn: initialization}
0 < r \leq \frac{\sigma}{2d}, & ~\alpha \in [1/10, 1/9] \cdot \frac{1}{\taubf(\theta)} \nonumber,  \\
v \in \Csig[\Xbf]^g, & ~\vol_0 \in [3/4,5/4] \cdot n(n-1) \int_{\Csig} \Pbb(B(x,r)) f(x) dx
\end{align}
where $\Csig[\Xbf]^g \subseteq \Csig[\Xbf]$ will be some large subset of $\Csig[\Xbf]$. In particular, letting $\vol_{n,r}(S) := \vol(S; G_{n,r})$ for $S \subseteq \Xbf$, we have $\vol_{n,r}(\Csig[\Xbf]^g) \geq \vol_{n,r}(\Csig[\Xbf])/2$.

@@

In practice it is clearly not feasible to set hyperparameters based on the underlying (unknown) density $f$. Typically, one tunes \pprspace over a range of hyperparameters and optimizes for some criterion such as minimum normalized cut; it is not obvious how this scheme would affect the performance of \pprspace in the density clustering context.

@@

The results of Section \ref{sec: analysis}, along with \eqref{eqn: graph_symmetric_set_difference_1}, give an upper bound on the volume of $\Cest \setminus \Csig[\Xbf]$ and $\Csig[\Xbf] \setminus \Cest$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference}
\vol_{n,r}(\Cest \setminus \Csig[\Xbf]) + \vol_{n,r}(\Csig[\Xbf] \setminus \Cest) \leq c \kappa(\Cset) \vol_{n,r}(\Csig[\Xbf]).
\end{equation}
To translate \eqref{eqn: graph_symmetric_set_difference} into meaningful bounds on the symmetric set difference $\Delta(\Csig[\Xbf], \Cest)$, we wish to preclude vertices $x \in \Xbf$ from having arbitrarily small degree, and so we make some regularity assumptions on $\mathcal{X} := \mathrm{supp}(f)$. Let $\nu$ denote the Lebesgue measure on $\Rd$, and $\nu_d := \nu(B)$ be the measure of the unit ball $B = B(0,1)$.
\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{5}
	\item 
	\label{asmp: valid_region}
	\emph{Regular support:} There exists some number $\lambda_{\min} > 0$ such that $\lambda_{\min} < f(x)$ for all $x \in \mathcal{X}$. Additionally, there exists some $c > 0$ such that for each $x \in \partial \mathcal{X}$, $\nu(B(x,r) \cap \mathcal{X}) \geq c \nu_d r^d$.
\end{enumerate}
Note that the latter condition in \ref{asmp: valid_region} will be satisfied if, for instance, the support $\mathcal{X}$ is a $\sigma$-expanded set.

\begin{theorem}
	\label{thm: misclassification_rate}
	Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a $\kappa$-well conditioned density cluster (with respect to some $\theta$), and additionally assume $f$ satisfies \ref{asmp: valid_region}. Then, there exists a universal constant $c_4 > 0$ such that with probability tending to $1$ as $n \to \infty$, 
	\begin{equation}
	\label{eqn: misclassification_rate_ub}
	\Delta(\Csig[\Xbf], \Cest) \leq c_4 \kappa(\Cset) \frac{\Lambda_{\sigma}}{\lambda_{\min}}.
	\end{equation}
\end{theorem}

The proof of Theorem \ref{thm: misclassification_rate}, along with all other proofs in this paper, can be found in the supplementary material. We observe that the symmetric set difference $\Delta(\Csig[\Xbf], \Cest)$  is proportional to the difficulty of the clustering problem, as measured by the condition number $\kappa(\Cset)$.

@@

We provide numerical experiments to investigate the tightness of our bounds on normalized cut and mixing time of $\Csig[\Xbf]$, and examine the performance of \pprspace on the ``two moons'' dataset. For space reasons, we defer details of the experimental settings to the supplement.

>>>>

\newcommand{\diam}{\rho}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
% \newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\Phi_{u}}
\newcommand{\Psibf}{\Psi_{u}}
\newcommand{\dist}{\mathrm{dist}}

@@

\begin{abstract}
\vskip 0.1 in
  Spectral clustering methods are a family of popular nonparametric clustering
  tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
  which extract clusters using locally-biased random walks around a user-specified
  seed node.  In contrast to existing results, we analyze PPR in a traditional
  statistical learning setup, where we obtain samples from an unknown
  distribution, and aim to identify connected regions of high-density (density
  clusters).  We prove that PPR, run on a neighborhood graph, extracts
  sufficiently salient density clusters, and provide empirical support for our theory.
\end{abstract}

@@

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (like
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of a local clustering algorithm on $\Xbf$ reveals
about the unknown density $f$.  In particular, we examine the ability of the \pprspace algorithm to recover \emph{density clusters} of $f$, which are defined as the
connected components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$
for some threshold $\lambda > 0$ (a central object of central interest in the
classical statistical literature on clustering, dating back to
\citet{hartigan1981}).

\paragraph{PPR on a neighborhood graph} We now describe the clustering algorithm that will be our focus for the rest of 
the paper. We start with the geometric graph that we form based on the samples 
$\Xbf$: for a radius $r > 0$, we consider the \emph{$r$-neighborhood graph} of 
$\Xbf$, denoted $G_{n,r}=(V,E)$, an unweighted, undirected graph with vertices
$V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i - x_j}
\leq  r$, where $\norm{\cdot}$ denotes Euclidean norm. We denote by $\Abf \in \Reals^{n \times n}$ the adjacency matrix, with entries $\Abf_{uv} = 1$ if and only if $(u,v) \in E$, and by $\Dbf$ the diagonal degree matrix, with $\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$.

Next, we define the PPR vector $\pbf = \pbf(v,\alpha;G_{n,r})$, with respect to  
a seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, to be
the solution of the following linear system:
\begin{equation}
\label{eqn: ppr_vector}
\pbf = \alpha \ebf{v} + (1 - \alpha) \pbf \Wbf,
\end{equation}
where $\Wbf = (\Ibf{} + \Dbf^{-1}\Abf)/2$ is the (lazy) random walk matrix over $G_{n,r}$ 
and $e_{v}$ denotes indicator vector for node $v$ (with a 1 in the $v$th
position and 0 elsewhere).

For a level $\beta > 0$ and a target volume $\vol_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \set{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}}.
\end{equation}
We need a metric to determine which sweep cut $S_{\beta}$ is the best cluster estimate. For a set $S \subseteq V$ with complement $S^c = V \setminus S$, we define the cut as $\cut(S;G_{n,r}) := \sum_{u \in S, v \in S^c} \Abf_{uv}$, the volume to be $\vol(S; G_{n,r}) := \sum_{u \in S} \Dbf_{uu}$, and the \emph{normalized cut} as
\begin{equation}
\label{eqn: normalized_cut}
\Phi(S; G_{n,r}) := \frac{\cut(S;G_{n,r})}{\min \set{\vol(S; G_{n,r}), \vol(S^c; G_{n,r})}}.
\end{equation}
Having computed sweep cuts $S_{\beta}$ over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}, we then output the cluster estimate $\widehat{C} = S_{\beta^*}$ which has minimum normalized cut $\Phi(S_{\beta^{\star}}; G_{n,r})$. For concreteness, we summarize this procedure in Algorithm \ref{alg: ppr}.

@@

\begin{definition}[Symmetric set difference]
	\label{def: symmetric_set_diff}
	For an estimator \smash{$\Cest \subseteq \Xbf$} and set
	$\mathcal{S} \subseteq \Reals^d$, the symmetric set difference of $\Cest$ and $\mathcal{S} \cap \Xbf$ is 
	\begin{equation}
	\label{eqn: misclassification_rate}
	\Delta(\Cest, \mathcal{S}) := \abs{\Cest \setminus \mathcal{S}[\Xbf] \cup \mathcal{S}[\Xbf] \setminus \Cest}.
	\end{equation}
\end{definition}

@@

\begin{enumerate}
	\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we introduce a set of natural geometric conditions, formalize a measure of difficulty based on these geometric conditions, and show that when properly initialized, the symmetric set difference of Algorithm \ref{alg: ppr} is upper bounded by this difficulty measure.

	\item We further show that if the density cluster $\Cset$ is particularly well-conditioned, Algorithm \ref{alg: ppr} consistently estimate a density cluster in the sense of \eqref{eqn: consistent_density_cluster_recovery}.

	\item In Section \ref{sec: analysis}, we detail some of the main technical machinery required to prove our main results, and expose the part various geometric quantities play in the ultimate difficulty of the clustering problem.

	\item In Section \ref{sec: experiments}, we empirically
	demonstrate the tightness of the bounds in Theorems \ref{thm: conductance_upper_bound} and \ref{thm: mixing_time_upper_bound}, and provide examples showing how violations of the geometric conditions we require manifestly
	impact density cluster recovery by \ppr.  
\end{enumerate}

@@

\paragraph{Related Work. }

In addition to the background given previously, a few related lines of work are worth
highlighting. Similar in spirit to our results are the works
\citep{shi2009,schiebinger2015}, who study the consistency of
spectral algorithms in recovering the latent labels in certain parametric and
nonparametric mixture models. These results focus on global rather than local
algorithms, and as such impose global rather than local conditions on the nature
of the density. Moreover, they do not in general guarantee recovery of density
clusters, which is the focus in our work. Perhaps most importantly, these works rely on general cluster saliency conditions, which depend implicitly on many distinct geometric aspects of the cluster $\Cset$ under consideration. We make this dependence explicit, and in so doing, expose the role each geometric condition plays in the clustering problem.

Additionally, we note that density clustering and level set estimation is a well-studied problem. \citep{polonik1995, rigollet2009} study density clustering under symmetric set difference, \citep{tsybakov1997, singh2009} prove minimax optimal level set estimators under Hausdorff loss and \citep{hartigan1981, chaudhuri2010} consider consistent estimation of the cluster tree, to note but a few works on the subject. Our goal is not to improve on these results, or offer yet another algorithm for level set estimation; indeed, seen as a density clustering algorithm, \pprspace has none of the optimality guarantees of the previous works. This is in fact a major point of our article: \pprspace can provably recover density clusters, but only under strong geometric conditions.

@@

As mentioned previously, successful recovery of a density cluster by \pprspace requires the density cluster to be geometrically well-conditioned. At a minimum, we wish to avoid sets $\Cset$ which contain arbitrarily thin bridges or spikes, and therefore as in \cite{chaudhuri2010} we introduce a buffer zone around $\Cset$. Let $B(x,r)$ be the closed ball of radius $r > 0$ centered at $x \in \Rd$. For a some $\lambda > 0$, consider a given cluster $\Cset \in \Cbb_f(\lambda)$. We denote the distance between $x$ and $\Cset$ as $\dist(x,\Cset) := \inf_{y \in \Cset} \norm{y - x}$, and for a given $\sigma > 0$, we refer to $\Csig := \set{x \in \Reals^d: \dist(x,\Cset) \leq \sigma}$ as the $\sigma$-expansion of $\Cset$. We now state our conditions with respect to $\Csig$, and provide some intuition afterwards.

@@

	\item 
	\label{asmp: low_noise_density}
	\emph{Low noise density:} There exists $\gamma,c_0 > 0$ such that for all $x
	\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
	$$
	\inf_{x' \in \Csig} f(x') - f(x) \geq  c_0 \dist(x, \Csig)^{\gamma},
	$$

@@

Thinking of $\Csig[\Xbf]$ as a subset of vertices in $G_{n,r}$, we would like $\Csig[\Xbf]$ to be internally well-connected, while being poorly connected to the rest of $\Xbf$. The cluster separation \ref{asmp: cluster_separation} and low noise density \ref{asmp: low_noise_density} conditions guarantee low connectivity between $\Csig[\Xbf]$ and $\Xbf \setminus \Csig[\Xbf]$ in $G_{n,r}$, whereas \ref{asmp: bounded_density} and \ref{asmp: embedding} ensure high connectivity within $\Csig[\Xbf]$. It may not be immediately obvious how \ref{asmp: embedding} contributes to geometric conditioning. For now, we observe merely that random walks will mix slowly over sets with large diameter, and comment on this condition in more detail in Section \ref{sec: analysis}. Finally, \ref{asmp: bounded_volume} is a relatively harmless technical condition, merely excluding the case where $\vol(\Csig[\Xbf]; G_{n,r}) > \vol(\Xbf; G_{n,r})/2$.

We can now formally define the condition number, $\kappa(\Cset)$, which reflects the difficulty of the local spectral clustering task. The smaller $\kappa(\Cset)$ is, the more success \pprspace will have in recovering $\Cset$. Let $\theta := (r, \sigma, \lambda, \lambda_{\sigma}, \Lambda_{\sigma}, \gamma, \diam, L)$ contain those geometric parameters detailed in \ref{asmp: bounded_density} - \ref{asmp: bounded_volume}.

\begin{definition}[Well-conditioned density clusters]
	For $\lambda > 0$ and $\Cset \in \Cbb_f(\lambda)$, let $\Cset$ satisfy \ref{asmp: bounded_density} - \ref{asmp: bounded_volume} for some $\theta$, Then, for universal constants $c_1, c_2, c_3 > 0$ to be specified later, we set
	\begin{equation}
	\label{eqn: condition_number_1}
	\Phibf(\theta) 
	:= c_1 r \frac{d}{\sigma} \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} - c_0 \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}},~ 
	\Psibf(\theta) := \Biggl(c_2 \frac{\Lambda_{\sigma}^4 d^3 D^2 L^2}{\lambda_{\sigma}^4 r^2} \log^2\left(\frac{1}{r}\right) + c_3 \Biggr)^{-1}
	\end{equation}
	and letting $\kappa(\Cset) := \frac{\Phibf(\theta)}{\Psibf(\theta)}$, we call $\Cset$ a \textrm{$\kappa$-well-conditioned density cluster}.
\end{definition}

At first glance \eqref{eqn: condition_number_1} may appear mysterious, but as will be shown in Section \ref{sec: analysis}, $\Phibf(\theta)$ and $\Psibf(\theta)$ are merely upper bounds on the normalized cut \eqref{eqn: normalized_cut} of $\Csig[\Xbf]$ in $G_{n,r}$, and the inverse of the mixing time (defined in Section \ref{sec: analysis} by \eqref{eqn: mixing_time}) of $\Csig[\Xbf]$ in $G_{n,r}$. In \cite{zhu2013}, building on the work of \cite{andersen2006} and others, it is shown that the ratio of normalized cut to inverse mixing time (or equivalently, the product of normalized cut and mixing time) is a fundamental quantity governing the clustering performance of \pprspace on a general graph. The condition number $\kappa(\Cset)$ is an asymptotic upper bound of this ratio for an empirical density cluster over the neighborhood graph $G_{n,r}$, and is therefore a natural criterion to measure difficulty of the density clustering task.

@@

As is typical in the local clustering literature, our algorithmic results will be stated with respect to specific choices or ranges of each of the user-specified parameters.

In particular, for a well-conditioned density cluster $\Cset$ (with respect to some $\theta$), we require
\begin{align}
\label{eqn: initialization}
r \leq \frac{\sigma}{2d}, & ~\alpha \in [1/10, 1/9] \cdot \Psibf(\theta) \nonumber,  \\
v \in \Csig[\Xbf]^g, & ~\vol_0 \in [3/4,5/4] \cdot n(n-1) \int_{\Csig} \Pbb(B(x,r)) f(x) dx
\end{align}
where $\Csig[\Xbf]^g \subseteq \Csig[\Xbf]$ will be some large subset of $\Csig[\Xbf]$. In particular, letting $\vol_{n,r}(S) := \vol(S; G_{n,r})$ for $S \subseteq \Xbf$, we have $\vol_{n,r}(\Csig[\Xbf]^g) \geq \vol_{n,r}(\Csig[\Xbf])/2$.

@@

In practice it is clearly not feasible to set hyperparameters based on the underlying (unknown) density $f$. Typically, one tunes \pprspace over a range of hyperparameters and optimizes for some criterion such as minimum normalized cut; it is unclear how this scheme would affect the performance of \pprspace in the density clustering context.

@@

Theorem 1 of \cite{zhu2013}, combined with the results of Section \ref{sec: analysis}, immediately implies a bound on the volume of $\Cest \setminus \Csig[\Xbf]$ (and likewise $\Csig[\Xbf] \setminus \Cest$),\footnote{For sequences $a_n$ and $b_n$, we write $a_n \lesssim b_n$ if there exists constant $c$ such that $a_n \leq c b_n$ for all $n$ sufficiently large. }
\begin{equation}
\label{eqn: graph_symmetric_set_difference}
\vol_{n,r}(\Cest \setminus \Csig[\Xbf]), \vol_{n,r}(\Csig[\Xbf] \setminus \Cest) \lesssim \kappa(\Cset) \vol_{n,r}(\Csig[\Xbf]).
\end{equation}
To translate \eqref{eqn: graph_symmetric_set_difference} into meaningful bounds on the symmetric set difference $\Delta(\Csig[\Xbf], \Cest)$, we wish to preclude vertices $x \in \Xbf$ from having arbitrarily small degree. To do so, we make some regularity assumptions on $\mathcal{X} := \mathrm{supp}(f)$. Let $\nu$ denote the Lebesgue measure on $\Rd$, and $\nu_d = \nu(B)$ be the measure of the unit ball $B = B(0,1)$.
\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{5}
	\item 
	\label{asmp: valid_region}
	\emph{Regular support:} There exists some number $\lambda_{\min} > 0$ such that $\lambda_{\min} < f(x)$ for all $x \in \mathcal{X}$. Additionally, there exists some $c > 0$ such that for each $x \in \partial \mathcal{X}$, $\nu(B(x,r) \cap \mathcal{X}) \geq c \nu_d r^d$.
\end{enumerate}
Note that the latter condition in $\ref{asmp: valid_region}$ will be satisfied if, for instance, the support $\mathcal{X}$ is a $\sigma$-expanded set.

\begin{theorem}
	\label{thm: misclassification_rate}
	Fix $\lambda > 0$, let $\Cset \in \Cbb_f(\lambda)$ be a $\kappa$-well conditioned density cluster (with respect to some $\theta$), and additionally assume $f$ satisfies \ref{asmp: valid_region}. Then, there exists a universal constant $c_4 > 0$ such that with probability tending to one as $n \to \infty$, 
	\begin{equation}
	\label{eqn: misclassification_rate_ub}
	\Delta(\Csig[\Xbf], \Cest) \leq c_4 \kappa(\Cset) \frac{\Lambda_{\sigma}}{\lambda_{\min}}.
	\end{equation}
\end{theorem}

The proof of Theorem \ref{thm: misclassification_rate}, along with all other proofs in this paper, can be found in the supplementary material. We observe that the symmetric set difference $\Delta(\Csig[\Xbf], \Cest)$  is proportional to the difficulty of the clustering problem, as measured by the condition number.

@@

We provide numerical experiments to investigate the tightness of our bounds in on normalized cut and mixing time of $\Csig[\Xbf]$, and examine the performance of \pprspace on the 'two moons' dataset. For space reasons, we defer details of the experimental settings to the supplement.
